Embedding Intuitionistic into Classical Logic

### 2023-03-10 LPAR/LPAR-EIICL.tex
Adressed Remarks

\begin{definition}\label{definition:skolemization}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to a set of free variables $Z$ by simultaneous induction:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$ and $(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi^S_{Z\cup\{x\}})$.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi^H_{Z\cup\{x\}})$.
=======
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi[a/x]^S_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi[a/x]^H_{Z\cup\{a\}}[x/a])$
 where $a$ is a new free variable.
>>>>>>> c8d6f4a7fa194cc08089d36f91f2f60b706bd539
	\end{itemize}
	Let $\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$ and $\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a QBF $\varphi^Q$ \todo{Ich würde die Formeln $\varphi^Q$ hier und in Section 6 aus Symmetriegründen auch $\varphi^\#$ nennen.} with $\mathcal |\varphi^Q|\in\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$ such that $\varphi$ is intuitionistically invalid if and only if $\varphi^Q$ is a satisfiable QBF.
\end{lemma}

@@

\begin{lemma}
	Suppose $(M, T)$ is a counter-model to $\varphi^{CH}$, then so is $(M_T, I_T)$.
\end{lemma}
\begin{proof}
	It follows directly from the definitions that $M_T, I_T\not\models P(s)$ and $M_T, I_T\models K(\varphi)$. It is easy to show that $M_T, I_T\models \psi$ for $\psi\in\mathcal R^{bS}$ from persistency. Let $\psi\in\mathcal X^{bS}$, we show $M_T, I_T[\psi_1\dots\psi_n/w] \models s\preceq w\to (w\preceq f_\psi(w)\to A(f_\psi(w)\to B(f_\psi(w)))\to C(w))$ for all $\psi_1\dots\psi_n\in M_T$. Let $\psi_1\dots\psi_n\in M_T$ and $m = g_{\psi_n}(\dots(g_{\psi_1}(s^I))$. If $M, I[m/w]\models C(w)$ we are done. Otherwise $M, I[m/w]\not\models s\preceq w$ or $M, I[m/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w)) \to B(f_\psi(w))$. The former is impossible since $g_\psi$ is increasing for all $\psi$. For the latter to hold we must have $M, I[m/w]\models w\preceq f_\psi(w)$, $M, I[m/w]\models A(f_\psi(w))$ and $M, I[m/w]\not\models B(f_\psi(w))$. If $\psi$ is fulfilled at $m$ we have $g_\psi(m) = m$, otherwise $g_\psi(m) = f_{[\psi]}^I(m)$. In any case $M, I'[\psi_1\dots\psi_n/w]\models A(f_\psi(w))$ and $M_T, I_T[\psi_1\dots\psi_n/w]\not\models B(f_\psi(w))$. It follows that $M_T, I_T[\psi_1\dots\psi_n/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w))$ and therefore $M_T, I_T\models \psi$.
=======
	1. If $C^I(m)$ is true then we are done due to persistency. Otherwise $C^I(m)$ is false.
	Because $(A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m)))\to C^I(m)$ holds, then $A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m))$ must be false.

	2. Let $n\succeq^I m$.
	If $C^I(n)$ is true, we are done.
	Otherwise $C^I(n)$ is false.
	Due to persistency $C^I(m)$ is also false.
	Because $\psi$ is fulfilled at $m$, $A^I(m)\to B^I(m)$ must be false, i.e. $A^I(m)$ is true and $B^I(m)$ is false. Then $A^I(n)$ and $A^I(f^I_{[\psi]}(n))$ are also true due to persistency.
	Because $(A^I(f^I_{[\psi]}(n))\to B^I(f^I_{[\psi]}(n)))\to C^I(n)$ holds, we now get that  $B^I(f^I_{[\psi]}(n))$ must be false.
	But then due to persistency so is $B^I(n)$ and we have that $\psi$ is fulfilled at $n$.
>>>>>>> c8d6f4a7fa194cc08089d36f91f2f60b706bd539
\end{proof}

@@

\begin{theorem}\label{thm:prop-translation}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implication $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $\mathcal R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to C^\lambda\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $P^\#$ \todo{Hier ist nur P statt mathcal P, was aus Symmetriegründen vielleicht hilfreich wäre, aber P wird auch schon für die finale Präposition verwendet, also wäre ein anderer Buchstabe wohl besser} as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$, encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results are stated in Figure \ref{fig:a}. 
In general, our approach falls short of the state of the art. 
=======
\vspace*{-.5cm}
Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results are stated in Figure \ref{fig:a}.
In general, our approach falls short of the state of the art.
>>>>>>> c8d6f4a7fa194cc08089d36f91f2f60b706bd539
However when grouping the benchmarks by $|\mathcal Q|$ and $|\mathcal X|$, see Figure~\ref{fig:b},
we observe good performance of our approach for $|\mathcal Q| = 0$ and $|\mathcal X| \leq 2$ resp. $|\mathcal X| \geq 6$.
Figure~\ref{fig:c} shows that our approach solved 52 problems not solved by the state-of-the-art provers. Notably, while there is big overlap in solved instances, ileanCoP and nanoCoP also have a significant symmetric difference.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.8\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3+	&33&18.9\%	&38&21.7\%		&70&40.0\%		&51&29.1\%	&175\\
		1-2&0-2	&84&29.3\%	&81&28.2\%		&227&79.1\%		&221&77.0\%	&287\\
		1-2&3+	&102&24.7\%	&98&27.3\%		&180&43.6\%		&157&39.1\%	&413\\
		3+&0+	&66&5.4\%	&63&5.2\%		&262&20.7\%		&270&22.2\%	&1216\\
	\end{tabular}
	\caption{Solved problems grouped by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}

@@

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
Suppose $\varphi$ is not intuitionistically valid.
Then, $(\bigwedge\mathcal R\wedge\bigwedge\mathcal X)\to P$ has an intuitionistic counter-model. By the previous section\todo{unklar was mit "previous section" gemeint ist, und was für ein model du genau meinst.} there exists a classical counter-model $\mathcal M$ for $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$.
We now define a QBF model iteratively.
For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal X$ such that $\psi_i$ is exactly the $\psi\in\mathcal X$ for which $X_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$  in $\mathcal M$.
Let the $X^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one of the $X^{n}_\psi$ is interpreted as true, then valid$(n)$ fails and the remaining propositions can be chosen arbitrarily. If $X^i_{\psi_n} = 1$ for some $i < n$, then valid$(n)$ also fails and the remaining propositions can be chosen arbitrarily.
So we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions.
Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$.
Continue this construction until $n  = |\mathcal X|$. Then from $\mathcal M$ being a counter-example to $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$ it directly follows that this interpretation satisfies $\varphi^Q$.

>>>>

\begin{definition}\label{definition:skolemization}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to a set of free variables $Z$ by simultaneous induction:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$ and $(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi^S_{Z\cup\{x\}})$.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi^H_{Z\cup\{x\}})$.
	\end{itemize}
	Let $\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$ and $\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a QBF $\varphi^Q$ with $\mathcal |\varphi^Q|\in\mathcal O(|\mathcal R|\cdot|\mathcal X| + |\mathcal X|^3)$ such that $\varphi$ is intuitionistically invalid if and only if $\varphi^Q$ is a satisfiable QBF.
\end{lemma}

@@

\begin{lemma}
	Suppose $\mathcal M = (M, T)$ is a counter-model to $\varphi^{CH}$, then so is $\mathcal M_T = (M_T, I_T)$.
\end{lemma}
\begin{proof}
	Let $v$ be a variable assignment. It follows directly from the definitions that $\mathcal M_T, v\not\models P(s)$ and $\mathcal M_T, v\models K(\varphi)$. It is easy to show that $\mathcal M_T, v\models \psi$ for $\psi\in\mathcal R^{bS}$ from persistency. Let $\psi\in\mathcal X^{bS}$, we show $\mathcal M_T, v[\psi_1\dots\psi_n/w] \models s\preceq w\to (w\preceq f_\psi(w)\to A(f_\psi(w)\to B(f_\psi(w)))\to C(w))$ for all $\psi_1\dots\psi_n\in M_T$. Let $\psi_1\dots\psi_n\in M_T$ and $m = g_{\psi_n}(\dots(g_{\psi_1}(s^I))$. If $\mathcal M, v[m/w]\models C(w)$ we are done. Otherwise $\mathcal M, v[m/w]\not\models s\preceq w$ or $\mathcal M, v[m/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w)) \to B(f_\psi(w))$. The former is impossible since $g_\psi$ is increasing for all $\psi$. For the latter to hold we must have $\mathcal M, v[m/w]\models w\preceq f_\psi(w)$, $\mathcal M, v[m/w]\models A(f_\psi(w))$ and $\mathcal M, v[m/w]\not\models B(f_\psi(w))$. If $\psi$ is fulfilled at $m$ we have $g_\psi(m) = m$, otherwise $g_\psi(m) = f_{[\psi]}^I(m)$. In any case $\mathcal M_T, v[\psi_1\dots\psi_n/w]\models A(f_\psi(w))$ and $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models B(f_\psi(w))$. It follows that $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w))$ and therefore $\mathcal M_T, v\models \psi$.
\end{proof}

@@

\begin{theorem}\label{thm:prop-translation}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implicational $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $\mathcal R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to C^\lambda\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $\mathcal P^\#$ as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$, encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge \mathcal P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results are stated in Figure \ref{fig:a}. 
In general, our approach falls short of the state of the art. 
However when grouping the benchmarks by $|\mathcal Q|$ and $|\mathcal X|$, see Figure~\ref{fig:b},
we observe good performance of our approach for $|\mathcal Q| = 0$ and $|\mathcal X| \leq 2$ resp. $|\mathcal X| \geq 6$.
Figure~\ref{fig:c} shows that our approach solved 52 problems not solved by the state-of-the-art provers. Notably, while there is big overlap in solved instances, ileanCoP and nanoCoP also have a significant symmetric difference.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.8\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3+	&33&18.9\%	&38&21.7\%		&70&40.0\%		&51&29.1\%	&175\\
		1-2&0-2	&84&29.3\%	&81&28.2\%		&227&79.1\%		&221&77.0\%	&287\\
		1-2&3+	&102&24.7\%	&98&27.3\%		&180&43.6\%		&157&39.1\%	&413\\
		3+&0+	&66&5.4\%	&63&5.2\%		&262&20.7\%		&270&22.2\%	&1216\\
	\end{tabular}
	\caption{Solved problems grouped by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}

@@

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
Suppose $\varphi$ is not intuitionistically valid.
Then, $(\bigwedge\mathcal R\wedge\bigwedge\mathcal X)\to P$ has an intuitionistic counter-model. By theorem~\ref{thm:prop-translation} there exists a classical counter-model $\mathcal M$ for $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$.
We now define a QBF model iteratively.
For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal X$ such that $\psi_i$ is exactly the $\psi\in\mathcal X$ for which $X_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$  in $\mathcal M$.
Let the $X^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one of the $X^{n}_\psi$ is interpreted as true, then valid$(n)$ fails and the remaining propositions can be chosen arbitrarily. If $X^i_{\psi_n} = 1$ for some $i < n$, then valid$(n)$ also fails and the remaining propositions can be chosen arbitrarily.
So we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions.
Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$.
Continue this construction until $n  = |\mathcal X|$. Then from $\mathcal M$ being a counter-example to $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$ it directly follows that this interpretation satisfies $\varphi^Q$.
